working title Compatibility system and stigma size are the main predictors of heterospecific pollen effect

### 2019-02-21 Manuscript_draft/Manuscript_draft.Rmd
intro changes

For the understanding at what level or intensity the interference of pollen can occur is important to consider the relatedness of the interacting species. Closely related species are more likely to have similar traits. The similarity in traits between closely related species can lead to higher chances of ovule usurpation/abortion and therefore studies predict a greater negative effect of closely related species (REFS). Few studies however, have focused on the impacts of heterospecific pollen on fitness of distantly related species Thomson et al. (1982) Galen and Gregory (1989) Neiland and Wilcock (1999). Yet, most insects and most stigmas have been found to carry multiple species of foreign pollen with little attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang and Huang (2013); also cite studies from pollen transfer networks here such as... ). The relative amounts of closely/far related pollen that arrives to a stigma gives a first snapshoot where pollen competition can occur and improves the understanding of how important the reduction of fitness can be from heterospecific pollen.

CLOSELY RELATED SPECIES STUDIES:  (Refs? Brown and Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang (2016))

>>>>

For the understanding at what level or intensity the interference of pollen can occur is important to consider the relatedness of the interacting species. Closely related species are more likely to have similar traits @arceo2016. The similarity in traits between closely related species can lead to higher chances of ovule usurpation/abortion (REF) and therefore studies predict and show a greater negative effect of closely related species (REFS). Few studies however, have focused on the impacts of heterospecific pollen on fitness of distantly related species Thomson et al. (1982) Galen and Gregory (1989) Neiland and Wilcock (1999). Yet, most insects and most stigmas have been found to carry multiple species of foreign pollen with little attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang and Huang (2013). The relative amounts of closely/far related pollen that arrives to a stigma gives a first snapshoot where pollen competition can occur and improves the understanding of how important the reduction of fitness can be from heterospecific pollen where closely related species cand lead to greater detrimental effects @arceo2016.

### 2019-02-21 Manuscript_draft/Manuscript_draft.tex
intro changes

Plant traits are crucial to understand heterospecific pollen effect but
the multifactorial nature of the traits that are involve in the
pollen-pistil interaction make difficult to unravel what are the main
traits in driving the effect. These traits can be seen from a male
perspective of both donor and recipient where pollen size, pollen
aperture number and pollen allelopathy are key components to understand
the outcome of foreign pollen arrival (REFS). In Ashman and Arceo-Gómez
(2013) small pollen is predicted to cause a greater fitness decrease,
although this can be true there are also other possibilities to consider
which can obscure a predictive framework like big pollen can clogg small
stigmas with fewer pollen grains, bigger stigmas are less likely to be
clogged by small pollen grains and bigger pollen can outcompete smaller
pollen grains due to faster pollen tube growth rate Williams and Rouse
(1990). Moreover, to understand the different mechanical or chemical
effects of pollen also the female traits of the pollen recipient have to
be considered, from the literature these main traits are: stigma size,
style length, number of ovules, incompatibility system and also a
structural trait such as flower morphology Montgomery and Rathcke
(2012); Ashman and Arceo-Gómez (2013); Tong and Huang (2016). For
example, greater stigmatic area is positively correlated with greater
heterospecific pollen deposition Montgomery and Rathcke (2012) and
therefore possibly with an increase in negative effect. For species that
are self-incompatible the barriers towards heterospecific pollen are
stronger than self-compatible species Ashman and Arceo-Gómez (2013).
Nonetheless, an effect of foreign pollen is a bit obscured by the
variability within species, however species that are strong selfers or
strong outcrossers have less variablity in mating systems and
predictions of effect could be more realistic (see figure 1 from
Whitehead et al. (2018)). Although past research has progress in the
understanding of what traits can mediate the effect as we have shown
here, there are multiple traits involved and multiple possible scenarios
still to be explored empirically for a full understanding of the
importance of heterospecific pollen effect in nature.

For the understanding at what level or intensity the interference of
pollen can occur is important to consider the relatedness of the
interacting species. Closely related species are more likely to have
similar traits (Refs? Brown and Mitchell (2001) Arceo-Gómez et al.
(2016) Tong and Huang (2016)). The similarity in traits between closely
related species can lead to higher chances of ovule usurpation and
therefore studies predict a greater negative effect of closely related
species. Few studies however, have focused on the impacts of
heterospecific pollen on fitness of distantly related species Thomson et
al. (1982) Galen and Gregory (1989) Neiland and Wilcock (1999). Yet,
most insects and most stigmas have been found to carry multiple species
of foreign pollen with little attention to degree of relatedness
(Arceo-Gómez and Ashman (2016); Fang and Huang (2013); also cite studies
from pollen transfer networks here such as\ldots{} ). Traditionally,
studies have overlook the effect of closely related species but in a
natural context both can occur and different degree of relatedness
should be considered.

@@

>>>>

Plant traits are crucial to understand heterospecific pollen effect but
the multifactorial nature of the traits that are involve in the
pollen-pistil interaction make difficult to unravel what are the main
traits in driving the effect. These traits can be seen from a male
perspective of both donor and recipient where pollen size, pollen
aperture number and pollen allelopathy are key components to understand
the outcome of foreign pollen arrival Murphy and Aarssen (1995); @Ashman
and Arceo-Gómez (2013). In Ashman and Arceo-Gómez (2013) small pollen is
predicted to cause a greater fitness decrease, although this can be true
there are also other possibilities to consider which can obscure a
predictive framework like big pollen can clogg small stigmas with fewer
pollen grains, bigger stigmas are less likely to be clogged by small
pollen grains and bigger pollen can outcompete smaller pollen grains due
to faster pollen tube growth rate Williams and Rouse (1990). Moreover,
to understand the different mechanical or chemical effects of pollen
also the female traits of the pollen recipient have to be considered,
from the literature these main traits are: stigma size, style length,
number of ovules, incompatibility system and also a structural trait
such as flower morphology Montgomery and Rathcke (2012); Ashman and
Arceo-Gómez (2013); Tong and Huang (2016). For example, greater
stigmatic area is positively correlated with greater heterospecific
pollen deposition Montgomery and Rathcke (2012) and therefore possibly
with an increase in negative effect. For species that are
self-incompatible the barriers towards heterospecific pollen are
stronger than self-compatible species Ashman and Arceo-Gómez (2013).
Nonetheless, an effect of foreign pollen is a bit obscured by the
variability within species, however species that are strong selfers or
strong outcrossers have less variablity in mating systems and
predictions of effect could be more realistic (see figure 1 from
Whitehead et al. (2018)). Although past research has progress in the
understanding of what traits can mediate the effect as we have shown
here, there are multiple traits involved and multiple possible scenarios
still to be explored empirically for a full understanding of the
importance of heterospecific pollen effect in nature.

For the understanding at what level or intensity the interference of
pollen can occur is important to consider the relatedness of the
interacting species. Closely related species are more likely to have
similar traits Arceo-Gómez and Ashman (2016). The similarity in traits
between closely related species can lead to higher chances of ovule
usurpation/abortion (REF) and therefore studies predict and show a
greater negative effect of closely related species (REFS). Few studies
however, have focused on the impacts of heterospecific pollen on fitness
of distantly related species Thomson et al. (1982) Galen and Gregory
(1989) Neiland and Wilcock (1999). Yet, most insects and most stigmas
have been found to carry multiple species of foreign pollen with little
attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang
and Huang (2013). The relative amounts of closely/far related pollen
that arrives to a stigma gives a first snapshoot where pollen
competition can occur and improves the understanding of how important
the reduction of fitness can be from heterospecific pollen where closely
related species cand lead to greater detrimental effects Arceo-Gómez and
Ashman (2016).

@@

\hypertarget{ref-murphy1995}{}
Murphy, S. D., and L. W. Aarssen. 1995. Reduced seed set in elytrigia
repens caused by allelopathic pollen from phleum pratense. Canadian
Journal of Botany 73:1417--1422.
